One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-05-06 doc/genesetdp.tex
copyright

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{Gustavo S. Jeuken}
\author[2]{Lukas K\"{a}ll}
\affil[1]{Science for Life Laboratory, School of
Engineering Sciences in Chemistry, Biotechnology and Health,
Royal Institute of Technology -- KTH, Box 1031, 17121 Solna, Sweden\\ \texttt{gustavo.jeuken@scilifelab.se}}
\affil[2]{Science for Life Laboratory, School of
Engineering Sciences in Chemistry, Biotechnology and Health,
Royal Institute of Technology -- KTH, Box 1031, 17121 Solna, Sweden\\ \texttt{lukas.kall@scilifelab.se}}
\authorrunning{G. S. Jeuken and L. K\"{a}ll}
\begin{document}

@@

Users of NEA typically want to test if there is a fuctional association between a query set of genes and a pathway\cite{alexeyenko2012network}. We argue that such test are much easier implemented as tests of the query set, i.e. is the query set a randomly selected set of genes, or is there an overrepresentation of well-connected genes in the query set?

@@

\subsection{Random sampling algorithm}

@@

\begin{lstlisting}[caption={
The central part of the dynamic programing algorithm for finding $N(s)$. The function takes the vector of links per gene, $k_a$, as well as the number of query genes, $Q$ as an input. The function depends on two additional functions {\tt find\_maxscore(k,Q)}, which calculates the maximal score a query of size $Q$ can obtain, and {\tt ncr(a,b)}, which calculates $ a \choose b $.
}
, label=lst:gensetdp,captionpos=t,float,abovecaptionskip=-\medskipamount]
def genesetdp(k,Q):
    max_s = find_maxscore(k,Q)
    N = np.zeros((max_s+1,Q+1),dtype='float_')
    N[0,0] = 1

    for a in range(len(k)):
        for s in range(max_s,-1,-1):
            for c in range(Q,-1,-1):
                for b in range(k[a],0,-1): # Stop at b=1
                    if c-b>=0 and s-a*b>=0:
                        N[s,c] += ncr(k[a], b) * N[s-a*b,c-b]
    return N[:,q]
\end{lstlisting}
There is a memory efficient implementation. We first note that

@@

As a representative example for we chose the ``Glycolysis/Gluconeogenesis'' pathway, as it contained a number of genes that coincided with the median number of genes of the pathways in the definition file.

\subsection{Availability}

Source code of an implementation of GeneSetDP and GeneSetMC  as well as to the experiments conducte here in, is available from \url{https://github.com/statisticalbiotechnology/genesetdp}.

@@

In order to test the statistical calibration of our method, we calculated $p$~values for 10000 random picks of gene sets from the investigated genome. As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed. To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration} for $10000$ randomly assembeled queries from the human genome. As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.

@@

\begin{figure}[htb]
  \begin{center}
    \begin{tabular}{c@{}c@{}c}
      \includegraphics[width=5cm]{figures/calibration.png} &
      \includegraphics[width=5cm]{figures/calibration.png} &
      \includegraphics[width=5cm]{figures/calibration.png} \\
      $Q=10$ & $Q=20$ & $Q=30$
    \end{tabular}
  \end{center}
  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway. To test the $p$~values uniformity we plotted them against their normalized rank. We also added the calibration curve of BinoX, evaluated on the same random sets.}
  \label{fig:calibration}
\end{figure}

>>>>

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{Gustavo S. Jeuken}
\author[2]{Lukas K\"{a}ll}
\affil[1]{Science for Life Laboratory, School of
Engineering Sciences in Chemistry, Biotechnology and Health,
Royal Institute of Technology -- KTH, Box 1031, 17121 Solna, Sweden\\ \texttt{gustavo.jeuken@scilifelab.se}}
\affil[2]{Science for Life Laboratory, School of
Engineering Sciences in Chemistry, Biotechnology and Health,
Royal Institute of Technology -- KTH, Box 1031, 17121 Solna, Sweden\\ \texttt{lukas.kall@scilifelab.se}}
\authorrunning{G. S. Jeuken and L. K\"{a}ll}

\Copyright{Gustavo S. Jeuken and Lukas K\"{a}ll}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\subjclass{G.3. Probabilistic algorithms (including Monte Carlo)}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation".
\keywords{Gene Set Enrichment Analysis; Null model; p value; Sampling}% mandatory: Please provide 1-5 keywords

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\EventEditors{John Q. Open and Joan R. Acces}
%\EventNoEds{2}
\EventLongTitle{18th Workshop on Algorithms in Bioinformatics (WABI 2018)}
\EventShortTitle{WABI 2018}
\EventAcronym{WABI}
\EventYear{2018}
\EventDate{20–22 August 2018}
\EventLocation{Helsinki, Finland}
%\EventLogo{}
%\SeriesVolume{42}
%\ArticleNo{23}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

@@

Users of NEA typically want to test if there is a functional association between a query set of genes and a pathway\cite{alexeyenko2012network}. We argue that such test are much easier implemented as tests of the query set, i.e. is the query set a randomly selected set of genes, or is there an overrepresentation of well-connected genes in the query set?

@@

\subsection*{Random sampling algorithm}

@@

\begin{lstlisting}[caption={
The central part of the dynamic programing algorithm for finding $N(s)$. The function takes the vector of links per gene, $k_a$, as well as the number of query genes, $Q$ as an input. The function depends on two additional functions {\tt find\_maxscore(k,Q)}, which calculates the maximal score a query of size $Q$ can obtain, and {\tt comb(a,b)}, which calculates $ a \choose b $.
}
, label=lst:gensetdp,captionpos=t,float,abovecaptionskip=-\medskipamount]
def genesetdp(k,Q):
    max_s = find_maxscore(k,Q)
    N = np.zeros((max_s+1,Q+1))
    N[0,0] = 1

    for a in range(len(k)):
        for s in range(max_s,-1,-1):
            for c in range(Q,-1,-1):
                for b in range(k[a],0,-1): # Stop at b=1
                    if c-b>=0 and s-a*b>=0:
                        N[s,c] += comb(k[a], b) * N[s-a*b,c-b]
    return N[:,q]
\end{lstlisting}
There is a memory efficient implementation. We first note that

@@

For the purpose of demonstrating our argument, we selected a representative example in the ``Glycolysis/Gluconeogenesis'' pathway, as it contained a number of genes that coincided with the median number of genes of the pathways in the definition file.

\subsection*{Availability}

Source code of an implementation of GeneSetDP and GeneSetMC, as well as code for generating the plots of this paper, is available from \url{https://github.com/statisticalbiotechnology/genesetdp}.

@@

In order to test the statistical calibration of our method, we calculated $p$~values for 10000 random picks of gene sets from the investigated genome. As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed. To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration} for $10000$ randomly assembled queries from the human genome. As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.

@@

\begin{figure}[htb]
  \begin{center}
    \begin{tabular}{c@{}c@{}c}
      \includegraphics[width=5cm]{figures/calibration.png} &
      \includegraphics[width=5cm]{figures/calibration.png} &
      \includegraphics[width=5cm]{figures/calibration.png} \\
      $Q=10$ & $Q=20$ & $Q=30$
    \end{tabular}
  \end{center}
  \caption{{\bf Calibration of GeneSetDP and GeneSetMC.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway. To test the $p$~values uniformity we plotted them against their normalized rank. We also added the calibration curve of BinoX, evaluated on the same random sets.}
  \label{fig:calibration}
\end{figure}
